Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-14 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
some changes in Sec. III

This theory belongs to the class of scalar-Gauss-Bonnet theories, which are
described by the action
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \, \mathcal{G}
    \right]\,,
\end{equation}
%
where $g = \textrm{det} \, g$ is the metric determinant, $R$ is the Ricci
scalar and $\varphi$ is a dynamical scalar field that couples to the
Gauss-Bonnet invariant
%
$\mathcal{G} =
R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
- 4 R^{\mu\nu}R_{\mu\nu}
+ R^2$.
%
$R_{\mu\nu\rho\sigma}$ and $R_{\mu\nu}$ are the Riemann and Ricci tensors respectively.
%
By itself, $\int \dV \mathcal{G}$ is a boundary term in four dimensions
and hence does not contribute to the field equations~\cite{Myers:1987yn}.
%
However,  when coupled to $\varphi$, it can contribute to the field equation
through the coupling function $f(\varphi)$. The strength of the coupling is
set by $\ell_{\rm GB}$, with dimensions of length.

@@

Having reviewed the modified gravity theories we will consider, we now present
the sequential building blocks for the waveform model we use to test these theories
against GW observations.
%
We start by reviewing the theory-agnostic parametrized ringdown
spin-expansion coefficients (ParSpec) \sout{expansion} \ag{framework} \agcomm{replaced `framework' to prevent duplication.  The original parspec authors call this a data-analysis framework, so I think it would be safe to use it. I have left the term ``ParSpec expansion" in other places, because ParSpec is always abbreviated in those places.} in Sec.~\ref{sec:review_parspec}.
%
Next, in Sec.~\ref{sec:review_pSEOB}, we review our baseline
parametrized inspiral-(plunge)-merger-ringdown (IMR) BBH waveform
model~\cite{Brito:2018rfr,Ghosh:2021mrv} and explain how we modify
it to include the ParSpec expansion.
%
Finally, in Sec.~\ref{sec:theory_specific_qnm}, we show how we can map
theory-specific QNM calculations in modified gravity theories onto the
theory-agnostic free coefficients in the ParSpec expansion.
%
Ultimately, this provides us with an IMR BBH waveform model, with the ringdown
portion of the model informed by QNM calculations in specific beyond-GR
theories.

@@

We start by reviewing the ParSpec
framework introduced in Ref.~\cite{Maselli:2019mjd}, but following
some of the notation used in Ref.~\cite{Carullo:2021dui}.
%
A general procedure to describe deviations to the QNM frequencies $\omega_{k}^{\rm Kerr}$ and
damping times $\tau_{k}^{\rm Kerr}$ of the GR Kerr BH is to write,
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{k} &= \omega^{\rm Kerr}_{k} \, (1 + \delta \omega_{k}), \\
\tau_{k}   &= \tau^{\rm Kerr}_{k}   \, (1 + \delta \tau_{k}),
\end{align}
\end{subequations}
%
where $\delta\omega_{k}$ and $\delta\tau_{k}$ are the fractional deformation parameters.
%
Here we use the subscript $k$ to represent the QNM labels $\{l,\, m,\, n\}$,
where $l$ and $m$ are the multipole indices, and $n$ the overtone number.
%
This type of parametrization was adopted, for instance, in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Carullo:2018sfu}.

\ag{The current LIGO-Virgo-Kagra tests of BH ringdown (see Sec.VII.A of~\cite{LIGOScientific:2020tif} or Sec.VIII.A of~\cite{LIGOScientific:2021sio}) take a restrictive theory-agnostic approach towards the inference of $\delta\omega_{k}$ and $\delta\tau_{k}$. These deviations are either assumed to occur identically across all observed sources or belong to a generic underlying Gaussian population. A more realistic scenario, however, is that these parameters depend on the source BH's mass and spin.}
\sout{A caveat of parametrizations of the form~\eqref{eq:general_deviation}, is that
$\delta\omega_{k}$ and $\delta\tau_{k}$ depend, in general, on the BH's source mass and spin.}
%
Ideally, one would like to explicitly reinstate \sout{the dependence the source-frame
BH mass and spin} \ag{this dependence} for it would
%
(i) introduce deformation parameters which can be determined, once and for
all, from a specific gravity theory (GR included), and
%
(ii) make it simpler to combine constraints coming from multiple \ag{(independently observed)} sources.

ParSpec is such an observable-based
\sout{In the ParSpec expansion this is accomplished by performing a} bivariate
expansion of Eq.~\eqref{eq:general_deviation} in the detector-frame BH's mass
$M$ and spin $\chi$, \sout{namely}
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{k} &= \frac{1}{M} \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \omega^{(n)}_{k} \, \left( 1 + \gamma \delta \omega^{(n)}_{k} \right), \\
\tau_{k}   &= M     \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \tau^{(n)}_{k}   \, \left( 1 + \gamma \delta \tau^{(n)}_{k} \right),
\end{align}
\end{subequations}
%
where $\omega_{k}^{(n)}$ and $\tau_{k}^{(n)}$ are dimensionless coefficients of the
expansion in spin for Kerr QNMs\footnote{\ag{From this point onwards, to prevent ambiguity,
we will use $n$ to denote the series expansion index, and not the spherical
harmonic overtone.}};
%
$\delta\omega_{k}^{(n)}$ and $\delta\tau_{k}^{(n)}$ are source-independent
dimensionless coefficients that characterize the corrections to the Kerr QNM at
each spin-order;
%
and $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence on a given modified gravity theory is contained in the
dimensionless parameter $\gamma$, given by
%
\begin{equation}
\gamma
%
= \left(\frac{\ell}{M_{\rm s}}\right)^{p}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on a parameter $\ell$, with dimensions of length (and sets the
length-scale at which beyond-GR modifications become important), and
the exponent $p$ which is related to how the beyond-GR modifications are
added to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made \sout{$\ell$} \ag{$\gamma$} dimensionless  by using the other
length-scale associated with the remnant black hole, i.e., its source-frame mass
$M_{\rm s}$\footnote{\ag{Throughout this paper, we maintain the convention of
using the subscript ``s" to denote source-frame quantities and plain baseline
symbols for detector-frame measurements.}} which we can also write in terms of the detector-frame mass $M$ through
the redshift $z$~\cite{Krolak:1987ofj}. \ag{Also, dividing by the factor $G/c^2$ allows us to
express $\ell$ in physically intuitive metric units.}

>>>>

This theory belongs to the class of scalar-Gauss-Bonnet theories, which are
described by the action
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \, \mathcal{G}
    \right]\,,
\end{equation}
%
where $g = \textrm{det}(g)$ is the metric determinant, $R$ is the Ricci
scalar and $\varphi$ is a dynamical scalar field that couples to the
Gauss-Bonnet invariant
%
$\mathcal{G} =
R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
- 4 R^{\mu\nu}R_{\mu\nu}
+ R^2$.
%
$R_{\mu\nu\rho\sigma}$ and $R_{\mu\nu}$ are the Riemann and Ricci tensors respectively.
%
By itself, $\int \dV \mathcal{G}$ is a boundary term in four dimensions
and hence does not contribute to the field equations~\cite{Myers:1987yn}.
%
However,  when coupled to $\varphi$, it can contribute to the field equation
through the coupling function $f(\varphi)$. The strength of the coupling is
set by $\ell_{\rm GB}$, with dimensions of length.

@@

\ab{Having reviewed the modified gravity theories we will consider, we now present
the sequential building blocks for the waveform model we use to test these theories
against GW observations.
%
We start by reviewing the parametrized ringdown spin-expansion coefficients (ParSpec) framework in Sec.~\ref{sec:review_parspec}.
%
Next, in Sec.~\ref{sec:review_pSEOB}, we review our baseline
parametrized inspiral(--plunge)-merger-ringdown (IMR) BBH waveform
model~\cite{Brito:2018rfr,Ghosh:2021mrv} and explain how we extend 
it to include the ParSpec.
%
Finally, in Sec.~\ref{sec:theory_specific_qnm}, we show how we can map
theory-specific QNM calculations in modified gravity theories onto the
free coefficients in the ParSpec framework.}
%
Ultimately, this provides us with an IMR BBH waveform model, with the ringdown
portion of the model informed by QNM calculations in specific beyond-GR
theories.

@@

\ab{A general procedure to describe deviations to the QNM frequencies $\omega_{\ell m n}^{\rm Kerr}$ and
damping times $\tau_{\ell m n}^{\rm Kerr}$ of the GR Kerr BH is to write,}
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{\ell m n} &= \omega^{\rm Kerr}_{\ell m n} \, (1 + \delta \omega_{\ell m n}), \\
\tau_{\ell m n }   &= \tau^{\rm Kerr}_{\ell m  n}   \, (1 + \delta \tau_{\ell m n }),
\end{align}
\end{subequations}
%
\ab{where $\delta\omega_{\ell m n}$ and $\delta\tau_{\ell m  n}$ are the fractional deformation parameters,
$l$ and $m$ are the multipole indices, and $n$ the overtone number.}
%
This type of parametrization was adopted, for instance, in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Brito:2018rfr,Carullo:2018sfu,Ghosh:2021mrv}.

The current LIGO-Virgo-Kagra tests of BH ringdown (see Sec.VII.A
of Ref.~\cite{LIGOScientific:2020tif} or Sec.VIII.A
of Ref.~\cite{LIGOScientific:2021sio}) take a restrictive theory-agnostic
approach towards the inference of $\delta\omega_{\ell m n}$ and
$\delta\tau_{\ell m n}$. These deviations are either assumed to occur
identically across all observed sources or belong to a generic
underlying Gaussian population. A more realistic scenario, however, is
that these parameters depend on the source BH's mass and spin.
%
Ideally, one would like to explicitly reinstate this dependence, by 
%
(i) introducing deformation parameters which can be determined, once and for
all, from a specific gravity theory (GR included), and
%
(ii) making it simpler to combine constraints coming from multiple (independently observed) sources.

\ab{The ParSpec framework was introduced in Ref.~\cite{Maselli:2019mjd}. It is} an observable-based
bivariate expansion of Eq.~\eqref{eq:general_deviation}, given by
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{\ell m n} &= \frac{1}{M_{\rm f}} \sum_{j = 0}^{N_{\rm max}} \, \chi_{\rm f}^{j} \omega^{(j)}_{\ell m n} \, \left( 1 + \gamma\, \delta \omega^{(j)}_{\ell m n} \right), \\
\tau_{\ell m n}   &= M_{\rm f}     \sum_{j = 0}^{N_{\rm max}} \, \chi_{\rm f}^{j} \tau^{(j)}_{\ell m n}   \, \left( 1 + \gamma\, \delta \tau^{(j)}_{\ell m n} \right),
\end{align}
\end{subequations}
%
\ab{where $M_{\rm f}$ and $\chi_{\rm f}$ are the detector-frame final mass and spin, respectively;}
$\omega_{\ell m n}^{(j)}$ and $\tau_{\ell m n}^{(j)}$ are dimensionless coefficients of the
expansion in spin for Kerr QNMs; $\delta\omega_{\ell m n}^{(j)}$ and $\delta\tau_{\ell m n}^{(j)}$ are source-independent
dimensionless coefficients that characterize the corrections to the Kerr QNM at
each spin-order, 
%
and $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence for a given modified gravity theory is contained in the
dimensionless parameter $\gamma$, which reads
%
\begin{equation}
\gamma
%
= \left(\frac{\ell}{M_{\rm f}}\right)^{p}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M_{\rm f}}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on \ab{the length-scale parameter $\ell$ (non-GR modifications will become
important at distances $\lesssim \ell$) and 
the exponent $p$ which is related to how the beyond-GR modifications are
added to the Einstein-Hilbert action.}
%
In Eq.~\eqref{eq:def_gamma} we made \sout{$\ell$} \ag{$\gamma$} dimensionless  by using the other
length-scale associated with the remnant black hole, i.e., its source-frame mass
$M_{\rm s}$\footnote{\ag{Throughout this paper, we maintain the convention of
using the subscript ``s" to denote source-frame quantities and plain baseline
symbols for detector-frame measurements.}} which we can also write in terms of the detector-frame mass $M$ through
the redshift $z$~\cite{Krolak:1987ofj}. \ag{Also, dividing by the factor $G/c^2$ allows us to
express $\ell$ in physically intuitive metric units.}
